---
title: "m2b tutorial"
author: "Laurent Dubroca, Andréa Thiébault"
date: "`r Sys.Date()`"
output: html_vignette
bibliography: '/home/moi/datahome/work/biblio/enfin/biblioloran.bib'
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

#Introduction

Animal behaviour, including social interactions, are fundamental to the field of
ecology. Whereas the direct observation of animal behaviour is often limited due
to logistical constraints, collection of movement data have been greatly
facilitated through the development of bio-logging. Animal movement data
obtained through tracking instrumentation may potentially constitute a relevant
proxy to infer animal behaviour. This is, however, based on the premise that a
range of movement patterns can be linked to specific behaviours.

Statistical learning constitutes a number of methods that can be used to
assess the link between given variables from a fully informed training
dataset and then predict the values on a non-informed variable. We chose the
random forest algorithm for its capacity to deal with imbalanced data
(particularly relevant for behavioural data), its high prediction accuracy
and its ease of implementationi (@breiman2001b, @chen2004). 
The strength of random forest partly relies
in its ability to handle a very large number of variables. Hence, our
methodology is based on the derivation of multiple predictor variables from
the movement data over various temporal scales, in order to capture as much
information as possible on the changes and variations of movement.

In this tutorial, the link between behaviour and movement parameters is build
using a random forest model based on a seabird track with behavioural
observation made by a video camera deployed on the individual. A new class named
`xytb` is implemented in order to provide to the user an object where data,
derived information, modelling and results are included in a single object.

#A new class

`xytb` is a S4 class built to give the whole information related to a track (the track itself, the differentiation of the track parameters, 
the observed behaviour, the model and the prediction) in a single object. 
`xyt` stands for the track information, and `b` for the behaviour.
8 slots contains these information, and different methods are available to build
the different part of the object. The rationale behind this object is the possibility 
to keep everything (data, model and prediction) in the same container, and by
extension the possibility (1) to keep track of how precisely the prediction were
made and (2) to exchange the analyses with different users easily.




```{r diag,warn=FALSE,cache=TRUE,echo=FALSE,fig.cap="Schematic of the work flow from fieldwork to the results. The legend gives the representation of the different objects. The arrows between the boxes represent the use of the package functions and methods: in blue the computation of the data, in red the modelling, in green the outputs and in pink the export to the ltraj format (adehabitat package). The dotted lines between the xytb class and the R objects describe the slot of the class."}
DiagrammeR::grViz("
	digraph rmarkdown {
	graph[center=true,ratio=auto,rankdir=TD,compound=true];

	fieldwork[label='Fieldwork',color=white];
	fieldwork->behaviour[lhead=cluster0];
	fieldwork->descdata[lhead=cluster0];
	fieldwork->track[lhead=cluster0];

	subgraph cluster0{
		track[label='track data'];
		behaviour[label='behavioural data'];
		descdata[label='meta data'];
		label = 'data';
	}
	descdata->desc[fontcolor=blue,color=blue];
	behaviour->b[fontcolor=blue,color=blue];
	track->xyt[fontcolor=blue,color=blue];

	subgraph cluster1 {
		xytb[label='class xytb',shape=diamond];
		xytb-> desc[type=tee,style=dotted,dir=none];
		xytb-> xyt[type=tee,style=dotted,dir=none];
		xytb-> b[type=tee,style=dotted,dir=none];
		xytb-> dxyt[type=tee,style=dotted,dir=none];
		xytb-> befdxyt[type=tee,style=dotted,dir=none];
		xytb-> model[type=tee,style=dotted,dir=none];
		xytb-> rfcv[type=tee,style=dotted,dir=none];
		xytb-> predb[type=tee,style=dotted,dir=none];

		xyt->dxyt->befdxyt[color=blue,style=dashed];
		b->model[color=red,style=dashed];
		dxyt->model[color=red,style=dashed];
		befdxyt->model[color=red,style=dashed];
		model->predb[color=red,style=dashed];
		rfcv->model[dir=both,color=red];
		desc[shape=diamond,label='@desc:\nshort\ndescription']
		xyt[shape=diamond,label='@xyt:\ntrack']
		b[shape=diamond,label='@b:\nbehaviour']
		dxyt[shape=diamond,label='@dxyt:\ntrack\nderivative']
		befdxyt[shape=diamond,label='@befdxyt:\n@dxyt\nshifted']
		model[shape=diamond,label='@model:\nrandom forest\nmodel']
		rfcv[shape=diamond,label='@rfcv:\ncross validation\nof @model']
		predb[shape=diamond,label='@predb:\nprediction of \n@b using @model']
		label = 'xytb object';
	}

	subgraph cluster3 {
		label='Results';
		Plots[shape=box];
		Tables[shape=box];
	}

	xytb->Plots[color=green];
	xytb->Tables[color=green];

	ltraj[label='ltraj object',shape=diamond];
	xytb->ltraj[color=pink,dir=both];

	subgraph cluster100 {
		label='Legend'
		out[label='Output',shape=box];
		R[label='R object',shape=diamond];
		slot[label='slot',shape=diamond];
		fun[label='Functions\n& Methods',color=white];
		leg1[label='xytb()',color=blue,fontcolor=blue];
		leg2[label='modelRF()',color=red,fontcolor=red];
		leg3[label='resRF()\nand resB()',color=green,fontcolor=green];
		leg4[label='xytb2ltraj()\nand ltraj2xytb()',color=pink,fontcolor=pink];
		fun->leg1[color=blue];
		fun->leg2[color=red];
		fun->leg3[color=green];
		fun->leg4[color=pink];
		R->slot[style=dotted,dir=none];
		file[label='Data'];
	}


}")


```

#Methods and function overview

The different analytical steps are summarized in 4 main functions and methods. 

##`xytb`
`xytb` is a class, but also a method. 4 distincts signature help the user to load 
tracks and behavioural information in the object. During this step, 
derivated information of the track are computed and stored in the slot `dxyt`
and `befdxyt`. The parameter `winsize` and `idquant` let the user to specify the
windows on which to compute statistical operator, and some of them (the
quantiles mainly).

##`modelRF`

This function compute a random forest model to predict the behavioural
observation against the movement information. 

##`resRF` and `resB`
These two functions compute and plot the results related to the model (`resRF`)
and the behaviours (prediction vs observation, `resB`).

##`xytb2ltraj` and `ltraj2xytb`

These functions import or export an `xytb` object to the `ltraj` class. This
class is linked to the `adehabitatLT` package and the numerous function
dedicated to trajectories analyses (see @calenge2008).


#An example

##Data

The data frame `track_CAGA_005` contains the information the track and the
behavioural observation made on a gannet. Behavioural state at `-1` are
unobserved behaviours. This parameters has to be passed to the functions during
the analysis (see functions' help).

```{r data1,warn=FALSE,cache=TRUE,echo=TRUE}
library(m2b)
str(track_CAGA_005)
```

Different methods are available to build a `xytb` object. Here, the track and
behavioural information are taken from the data.frame, and derived information
related to the track are computed in the same time. The derived information are
computed on the moving windows of 3, 5, 7, 9, 11, 13 and 15 locations (the
`winsize` parameters), and with the standard statistical operators (mean, standard deviation
and median absolute deviation) available in the methods, the quantile at 0, 25,
50, 75 and 100\% are added (the `idquant` parameter). Then these values are
shifted back in time at 5, 10 and 15 points backward (the `move` parameter), in
order to build a dataset of predictor back in time. This is useful if the user
is interested to investigate the link of the observed behaviour with previous
movement events. The rationale behind this operation is based on the fact that
some change in movement can be triggered by behavioural observation made
afterward by the scientist but sooner by the animal.   

```{r data2,warn=FALSE,cache=TRUE,echo=TRUE}
library(m2b)
str(track_CAGA_005)
#convert to xybt object with computation of windows operators and some quantiles
xytb<-xytb(track_CAGA_005,desc="example track",
	 winsize=seq(3,15,2),idquant=seq(0,1,.25),move=c(5,10,15))
#a simple plot method
plot(xytb)
```

#Environnement

##Modelling

To build a random forest predicting the behavioural states based on the movement
information, the function `modelRF` is used. It's a simple wrapper calling the 
`randomForest` function of the `randomForest` package, using the behavioural
observation as response, and movement information as predictors.
Some diagnosis are available to check the results using the `resRF` function,
but the function `extractRF` made the model available in the `randomForest`
format, and let the use of the other function of the `randomForest` package.

```{r model1,warn=FALSE,cache=TRUE,echo=TRUE}
#a model (the function modelRF update the model inside the xytb object)
xytb<-modelRF(xytb,type="actual",ntree=501,mtry=15)
```

```{r model2,warn=FALSE,cache=TRUE,echo=TRUE}
resRF(xytb)
```

##Results

The results regarding the behavioural states predicted vs the one observed are
illustrated thanks to the `resB` functions.

```{r res1,warn=FALSE,cache=TRUE,echo=TRUE}
resB(xytb,"time",nob="-1")
resB(xytb,"space",nob="-1")
```

#Bibliography
